---
title: "Class 09: Candy Mini Project"
author: "Aarav Prasad (PID: A17440940)"
format: pdf
toc: true
---

## Background

Today we are taking a small detour to analyze a fun dataset that we have more intrinsic insight into. We will be using principal component analysis.Building on your work with PCA in the previous labs (UK Foods, RNA-seq and beast cancer biopsy analysis), this mini-project gives you the opportunity to apply your data analysis skills to a new, tastier dataset.


## Importing candy data

```{r}
candy_file <- "candy-data.txt"

candy = read.csv(candy_file, row.names = 1)
head(candy)
```

The dataset includes all sorts of information about different kinds of candy. For example, is a candy chocolaty? Does it have nougat? How does its cost compare to other candies? How many people prefer one candy over another?

>Q1: How many different candy types are in this dataset?

```{r}
nrow(candy)
```


>Q2: How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```


One of the most interesting variables in the dataset is `winpercent`. For a given candy this value is the percentage of people who prefer this candy over another randomly chosen candy from the dataset (what 538 term a matchup). Higher values indicate a more popular candy.

```{r}
candy["Twix", ]$winpercent
```

>Q3: What is your favorite candy (other than Twix) in the dataset and what is it’s `winpercent` value?

```{r}
candy["3 Musketeers", ]$winpercent
```

>Q4: What is the winpercent value for “Kit Kat”?

```{r}
library(dplyr)

candy |> 
  filter(row.names(candy)=="Kit Kat") |> 
  select(winpercent)
```

>Q5:  What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

**Side-note**: the `skimr::skim()` function

There is a useful `skim()` function in the **skimr** package that can help give you a quick overview of a given dataset. Let’s install this package and try it on our candy data.

```{r}
library("skimr")
skim(candy)
```

>Q6: Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

`winpercent` is on a different scale; it has a mean of 50 and a standard deviation of 15, while the rest of the columns are between 0 and 1.


>Q7: What do you think a zero and one represent for the `candy$chocolate` column?

A zero represents not containing chocolate, a one represents containing chocolate.


## Exploratory analysis

A good place to start any exploratory analysis is with a histogram. You can do this most easily with the base R function `hist()`. Alternatively, you can use `ggplot()` with `geom_hist()`. Either works well in this case and (as always) its your choice.

>Q8: Plot a histogram of `winpercent` values using both base R and ggplot2.


**Base R**

```{r}
hist(candy$winpercent)
```

**ggplot2**

```{r}
library("ggplot2")

ggplot(candy, aes(x = winpercent)) + geom_histogram(bins = 20)
```

>Q9: Is the distribution of winpercent values symmetrical?

The distribution is not symmetrical, there seems to be a higher concentration of winpercent values near 40 percent, skewed.

>Q10: Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```

The median is a better measure of centrality, and it is below 50%. 

>Q11: On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
mean(candy$winpercent[as.logical(candy$chocolate)])
```

```{r}
mean(candy$winpercent[as.logical(candy$fruity)])
```

Chocolate is higher ranked on average.

>Q12: Is this difference statistically significant?

```{r}
t.test(candy$winpercent[as.logical(candy$chocolate)], candy$winpercent[as.logical(candy$fruity)] )
```
It is significant, the p value is < 0.05.


## Overall Candy Rankings

Let’s use the base R `order()` function together with `head()` to sort the whole dataset by a variable/column of interest. Or if you have been getting into the tidyverse and the dplyr package you can use the `arrange()` function together with `head()` to do the same thing and answer the following questions:


>Q13: What are the five least liked candy types in this set?

```{r}
candy |>
  arrange(winpercent) |>
  head(5)
```


>Q14: What are the top 5 all time favorite candy types out of this set?


```{r}
candy |> 
  arrange(desc(winpercent)) |>
  head(5)
```


>Q15: Make a first barplot of candy ranking based on `winpercent` values

```{r, fig.height= 15, fig.width= 10}
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```


>Q16: This is quite ugly, use the `reorder()` function to get the bars sorted by `winpercent`

```{r, fig.height= 15, fig.width= 10}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col() 
```


**Time to add some useful color**

Let’s setup a color vector (that signifies candy type) that we can then use for some future plots. We start by making a vector of all black values (one for each candy). Then we overwrite chocolate (for chocolate candy), brown (for candy bars) and red (for fruity candy) values.

```{r}
my_cols=rep("black", nrow(candy))

my_cols[as.logical(candy$chocolate)] = "chocolate"

my_cols[as.logical(candy$bar)] = "brown"

my_cols[as.logical(candy$fruity)] = "pink"
```


Now let’s try our barplot with these colors. Note that we use `fill=my_cols` for `geom_col()`. Experiment to see what happens if you use `col=mycols`.

```{r, fig.height= 15, fig.width= 10}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) +
  ylab("")
```

>Q17: What is the worst ranked chocolate candy?

Sixlets

>Q18: What is the best ranked fruity candy?

Starburst


## Taking a look at pricepercent

What about value for money? What is the best candy for the least money? One way to get at this would be to make a plot of `winpercent` vs the `pricepercent` variable. The `pricepercent` variable records the percentile rank of the candy’s price against all the other candies in the dataset. Lower values are less expensive and higher values are more expensive.

```{r}
library(ggrepel)

# How about a plot of win vs price
ggplot(candy) +
  aes(x=winpercent, y=pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)

```

>Q19: Which candy type is the highest ranked in terms of `winpercent` for the least money - i.e. offers the most bang for your buck?

```{r}
candy$value = candy$winpercent / candy$pricepercent
candy[which.max(candy$value), ]
```

>Q20: What are the top 5 most expensive candy types in the dataset and of these which is the least popular?


```{r}
candy[order(candy$pricepercent, decreasing = TRUE),] [1:5,]
```
Nik L Nip is the least popular

## Exploring the correlation structure

We can calculate the pair-wise correlation of all our columns

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

>Q22: Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity are strongly anti-correlated. Bar and pluribus are also anti correlated.

>Q23: Use your corrplot result to make a prediction: which variables do you expect will have the largest contributions (i.e. loadings) to PC1 (i.e., drive the most separation between candies along the first principal component)?

 `Chocolate` and `fruity` will contribute most to PC1 because they are strongly negatively correlated and create a major split in the data. `Winpercent` and `pricepercent` also contribute since they have larger variation and help capture differences in popularity and cost. Together, these variables drive the main separation between candies.



## Principal Component Analysis

In this case we want to be sure to set `scale=TRUE` because we have one variable `winpercent` that is on an very different scale than all others and would otherwise dominate our PCA results.

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```
Now we can plot our main PCA score plot of PC1 vs PC2.

```{r}
plot(pca$x[,1:2])
```

We can change the plotting character and add some color:

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```
We can make a much nicer plot with the ggplot2 package but it is important to note that ggplot works best when you supply an input data.frame that includes a separate column for each of the aesthetics you would like displayed in your final plot.

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
  aes(x = PC1, y = PC2, 
      size = winpercent/100,  
      text = rownames(my_data),
      label = rownames(my_data)) +
  geom_point(col = my_cols)

p
```

Again we can use the ggrepel package and the function `ggrepel::geom_text_repel()` to label up the plot with non overlapping candy names like. We will also add a title and subtitle like so:

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

If you want to see more candy labels you can change the `max.overlaps` value to allow more overlapping labels or pass the `ggplot` object p to plotly like so to generate an interactive plot that you can mouse over to see labels:

```{r}
library(plotly)
```

```{r}
ggplotly(p)
```

Let’s finish by taking a quick look at PCA our loadings. Do these make sense to you? Notice the opposite effects of `chocolate` and `fruity` and the similar effects of `chocolate` and `bar` (i.e. we already know they are correlated).

```{r}
ggplot(pca$rotation) +
  aes(x = PC1, y = reorder(rownames(pca$rotation), PC1)) +
  geom_col()
```

>Q24:  Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

The variables picked up strongly in the positive direction of PC1 are `fruity`, `pluribus`, and `hard`. These make sense because they represent a group of non-chocolate, often bagged or hard candies, which contrasts with chocolate-based candies.

We previously saw this relationship in the correlation plot, where `fruity` and `chocolate` were strongly negatively correlated, showing the same separation that PC1 is capturing.


## Summary

The PCA loadings revealed that the primary axis of variation separates chocolate-based candies from `fruity` ones - the same pattern we observed in the correlation matrix, now visualized in a single, interpretable plot.

>Q25: Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

A “winning” candy is typically chocolate-based, often a bar, and not `fruity`, with higher winpercent and reasonable price. Visualizations showed chocolate candies clustering at higher popularity, while the correlation plot highlighted the strong negative relationship between `chocolate` and `fruity`. PCA confirmed this pattern by separating candies along PC1 based on `chocolate` vs `fruity`, reinforcing that candy type is the main driver of preference.





