---
title: "Class 08: Breast Cancer Mini Project"
author: "Aarav Prasad (PID: A17440940)"
format: pdf
toc: true
---

## Background

The goal of this mini-project is for you to explore a complete analysis using the unsupervised learning techniques covered in class

Today we will analyze a biopsy data-set from fine needle aspiration (FNA) of a breast mass.

## Data Import

The data is made available as a CSV file for download.

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names=1)

```

```{r}
head(wisc.df, 3)
```

Make sure we remove or exclude the `diagnosis` column from the data-set that we use for further analysis - this is the expert diagnosis as either M or B:

```{r}
wisc.data <- wisc.df[,-1]
diagnosis <- as.factor(wisc.df$diagnosis)
```


## Exploratory Data Analysis



>Q1: How many observations are in this dataset?

```{r}
nrow(wisc.data)
```

569 patients (observations)


>Q2: How many of the observations have a malignant diagnosis?

```{r}
table(wisc.df$diagnosis)
```

212 malignant


>Q3:How many variables/features in the data are suffixed with `_mean`?


We can use the `grep()` function to help us
```{r}
length(grep("_mean", colnames(wisc.data)))
```


## Principal Component Analysis (PCA)

We need to scale our data before PCA with the `scale=TRUE` argument to `prcomp()`. 

```{r}
wisc.pr <- prcomp(wisc.data, scale = TRUE)
summary(wisc.pr)
```

>Q4: From your results, what proportion of the original variance is captured by the first principal component (PC1)?

 0.4427 (44.27%)

>Q5: How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

At least 3 PCs

>Q6: How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

At least 7 PCs




# Interpreting PCA Results

Now you will use some visualizations to better understand your PCA model. A common visualization for PCA results is the so-called biplot.

```{r}
biplot(wisc.pr)
```

>Q7: What stands out to you about this plot? Is it easy or difficult to understand? Why?
 
 This plot is very difficult to understand because there are way too many data points/values to visualize on the graph, everything just blends together. The column names are red, but the rest of the numerical values all blend together.


Lets see the "PC score plot"

```{r}
library(ggplot2)

ggplot(wisc.pr$x) +
  aes(PC1,PC2, col = diagnosis) +
  geom_point()
```

>Q8:  Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

Both these plots show that PC1 captures a separation of malignant and benign samples.

```{r}

ggplot(wisc.pr$x) +
  aes(PC1, PC3, col=diagnosis) +
  geom_point()
```


# Variance Explained

A scree plot shows how much variance each PC captures. We typically look for an “elbow” — a point where adding more PCs gives diminishing returns.

Calculate the variance of each principal component by squaring the `sdev` component of `wisc.pr` (i.e. `wisc.pr$sdev^2`). Save the result as an object called `pr.var`.

```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

Calculate the variance explained by each principal component by dividing by the total variance explained of all principal components.

```{r}
# Variance explained by each principal component: pve
pve <- pr.var / sum(pr.var)

# Plot variance explained for each principal component
plot(c(1,pve), xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Percent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

## Communicating PCA Results

The loading vector (`wisc.pr$rotation`) tells us which original measurements contribute most to each PC.

>Q9: For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

```{r}
wisc.pr$rotation["concave.points_mean", 1]
sort(abs(wisc.pr$rotation[,1]))
```
The component of the loading vector is -2.608538, there aren't any features with larger contributions 


## Hierarchical clustering

As part of the preparation for hierarchical clustering, the distance between all pairs of observations needs to be calculated. This “distance matrix” will be the input for the `hclust()` function.



First scale the `wisc.data` data and assign the result to `data.scaled`.

```{r}
wisc.hclust <- hclust(dist(scale(wisc.data)))
plot(wisc.hclust)
```

>Q10: Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(h = 19, col = "red", lty = 2)
```
The clutstering model has 4 clusters at h = 19.

Use `cutree()` to cut the tree so that it has 4 clusters. Assign the output to the variable `wisc.hclust.clusters`.

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, h = 19)
```

```{r}
table(wisc.hclust.clusters, diagnosis)
```
As we discussed in our last class videos there are number of different “methods” we can use to combine points during the hierarchical clustering procedure. These include "single", "complete", "average" and (my favorite) "ward.D2"

>Q12: Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

I prefer **ward.D2** because it produces more compact and well-separated clusters by minimizing within-cluster variance. This makes the dendrogram easier to interpret compared to methods like single linkage, which can create chain-like clusters.



## Combining Methods

```{r}
d <- dist(wisc.pr$x[,1:4])
wisc.pr.hclust <-hclust(d, method = "ward.D2")
plot(wisc.pr.hclust)
abline(h=70, col = "red")
```

```{r}
grps <- cutree(wisc.pr.hclust, h=70)
table(grps)
```


How does this clustering `grps` correspond to the expert `diagnosis`? 

```{r}
table(diagnosis)
```


```{r}
table(diagnosis, grps)
```

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC2) +
  geom_point(col=grps)
```

```{r}
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k = 2)
```

Using `table()`, compare the results from your new hierarchical clustering model with the actual diagnoses.

>Q13: How well does the newly created hclust model with two clusters separate out the two “M” and “B” diagnoses?

```{r}
table(wisc.pr.hclust.clusters, diagnosis)
```
The 2-cluster model separates the data fairly well: one cluster is mostly malignant (165 M vs 6 B), and the other is mostly benign (351 B vs 47 M), though some misclassification remains.


>Q14: How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.hclust.clusters and wisc.pr.hclust.clusters) with the vector containing the actual diagnoses.

```{r}
table(wisc.hclust.clusters, diagnosis)
```
Without PCA, the clustering is less clean because the diagnoses are spread across multiple clusters instead of being clearly separated into two groups. Even if one cluster has fewer errors (e.g., 40 M), the overall separation is worse than the PCA-based 2-cluster model




##Prediction

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```
```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

>Q16: Which of these new patients should we prioritize for follow up based on your results?
 
 Patient 2, they are in the "malignant" group. 








