---
title: "Class 12 HW: Population Scale Analysis"
author: "Aarav Prasad (PID: A17440940)"
format: pdf
---

## Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3
expression.

>Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```
```{r}
table(expr$geno)
```

```{r}
b <- boxplot(exp ~ geno, data = expr)
b$stats[3, ]
```
>Q14:  Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
library(ggplot2)

```

```{r}
ggplot(expr) + aes(geno, exp, fill=geno) +
  geom_boxplot(notch=TRUE)
```
Based on the boxplot, ORMDL3 expression appears to differ by genotype. A/A individuals show the highest expression, G/G individuals show the lowest expression, and A/G individuals fall in between. This suggests rs8067378 may affect ORMDL3 expression, with the G allele associated with lower expression.
